\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm,bm}
\usepackage[margin=1in]{geometry}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}

\title{Is the arithmetic mean the optimal cluster representative?}
\author{}
\date{}

\begin{document}
\maketitle

\section{Setup}

Fix a cluster $C$ with $n=|C|$ keys $k_i$ and values $v_i$. For a query $q$ write the
scaled scores $s_i = q^\top k_i/\sqrt{d}$ and the within-cluster softmax weights and
value
\[
  w_i(q) = \frac{e^{s_i}}{\sum_{j\in C} e^{s_j}}, \qquad
  u_C(q) = \sum_{i\in C} w_i(q)\, v_i .
\]
The exact contribution of $C$ to attention is its mass $Z_C(q)=\sum_i e^{s_i}$ and its
value $u_C(q)$. The method stores a query-independent representative $(k^\ast,v^\ast)$ and
approximates
\[
  \hat Z_C(q) = n\, e^{q^\top k^\ast/\sqrt d}, \qquad \hat u_C = v^\ast,
\]
so that $\hat o = \sum_C \hat p_C v^\ast_C$ with $\hat p_C \propto \hat Z_C$. The current
choice is $k^\ast=\bar k=\frac1n\sum_i k_i$, $v^\ast=\bar v=\frac1n\sum_i v_i$. We ask,
for the \emph{query distribution $\mathcal Q$ that actually compresses $C$}, whether these
minimize the true output error, and how the certificate changes off the mean.

Let $\delta_C(q) = \max_i |s_i - q^\top\bar k/\sqrt d|$ and $B_C=\max_i\|v_i\|$. The
current certificate for the value channel is $\|\bar v - u_C\|\le 2B_C\tanh(\delta_C/2)$
and for the mass channel $2B(\cosh\delta_C-1)/S_\delta$.

\section{The value representative}

\begin{proposition}[MMSE storable value]\label{prop:value}
Among query-independent $v^\ast$, the minimiser of $\mathbb E_{q\sim\mathcal Q}\|v^\ast-u_C(q)\|_2^2$ is
\[
  \boxed{\;v^\ast_C = \mathbb E_{q\sim\mathcal Q}[u_C(q)] = \sum_{i\in C}\bar w_i\, v_i,\qquad
  \bar w_i = \mathbb E_{q\sim\mathcal Q}[w_i(q)]\;}
\]
a convex combination ($\bar w_i\ge0$, $\sum_i\bar w_i=1$), hence one stored vector. The
uniform mean $\bar v$ is optimal iff $\bar w_i\equiv 1/n$.
\end{proposition}
\begin{proof}
For a random vector $U=u_C(q)$ the constant minimising $\mathbb E\|c-U\|^2$ is
$c=\mathbb E U$; here $\mathbb E U=\sum_i \mathbb E[w_i]\,v_i$ by linearity. Each $w_i(q)\in[0,1]$
and $\sum_i w_i=1$ pointwise, so $\bar w$ is a probability vector.
\end{proof}

Thus $\bar v$ is optimal only under a symmetric query law; when the compressing queries
attend some keys more than others on average ($\bar w\neq\text{uniform}$), the
attention-weighted centroid strictly reduces the expected value error. The irreducible
floor is the query variance $\mathbb E\|u_C-v^\ast\|^2 = \mathbb E\|u_C\|^2-\|v^\ast\|^2$;
the oracle $u_C(q)$ attains zero and is the ceiling.

\begin{corollary}[Certificate: mean is worst-case optimal, $v^\ast$ is average-case optimal]
\label{cor:value-cert}
For any storable convex weight $a$ (so $v^\ast=\sum_i a_i v_i$),
\[
  \|v^\ast-u_C\| = \Big\|\sum_i (a_i-w_i)(v_i-c)\Big\| \le B_C^{(c)}\sum_i|a_i-w_i|
    = 2B_C^{(c)}\,\mathrm{TV}(a,w),
\]
with $B_C^{(c)}=\max_i\|v_i-c\|$. The uniform $a=1/n$ is the Chebyshev centre of the set
of $\delta$-reachable weights $\{w(q):|s_i-\bar s|\le\delta\}$, hence minimises
$\max_q\mathrm{TV}(a,w)$ and recovers $2B_C\tanh(\delta_C/2)$. Any $\bar w\neq$ uniform has
$\max_q\mathrm{TV}(\bar w,w)\le 2\tanh(\delta_C/2)$ (triangle inequality), i.e. the
certificate keeps the same TV form with a worst-case constant at most doubled, while the
\emph{expected} error strictly drops.
\end{corollary}

This is the precise sense in which ``the bound changes'': $\bar v$ optimises the
worst-case certificate, $v^\ast$ optimises the actual (average) error. Under the stated
objective (actual error, certificate a valid/computable constraint) $v^\ast$ dominates.

\section{The key representative}

\begin{proposition}[Mass bias and loss of the $\cosh$ certificate]\label{prop:key}
For $k^\ast=\bar k+\Delta$ and $\mu(q)=q^\top\Delta/\sqrt d$,
\[
  \frac{\hat Z_C(k^\ast)}{Z_C} = e^{\mu(q)}\,\frac{\hat Z_C(\bar k)}{Z_C},\qquad
  \frac{\hat Z_C(\bar k)}{Z_C}=\frac{1}{\tfrac1n\sum_i e^{x_i}},\ \ x_i=s_i-q^\top\bar k/\sqrt d .
\]
At $k^\ast=\bar k$ the deviations are mean-zero ($\frac1n\sum_i x_i=0$), which is exactly
the hypothesis of $\frac1n\sum_i e^{x_i}\le\cosh\delta_C$. Any $\Delta\neq0$ (i) makes the
per-query mass estimate first-order biased by $e^{\mu(q)}$, and (ii) violates mean-zero, so
the mass certificate must be re-derived as $\frac1n\sum_i e^{x_i}\le e^{\mu}\cosh\delta_C'$,
generally looser.
\end{proposition}
\begin{proof}
$\hat Z_C(k^\ast)=n e^{q^\top k^\ast/\sqrt d}=e^{\mu} n e^{q^\top\bar k/\sqrt d}
=e^{\mu}\hat Z_C(\bar k)$, and $Z_C=e^{q^\top\bar k/\sqrt d}\sum_i e^{x_i}$. The bound
$\frac1n\sum e^{x_i}\le\cosh\delta$ under $|x_i|\le\delta,\ \sum x_i=0$ is tight at the
two-point law $\pm\delta$; dropping $\sum x_i=0$ shifts the extremiser and adds the $e^\mu$
factor.
\end{proof}

The $\Delta$ minimising the mean log-mass error $\mathbb E_{\mathcal Q}\big[\log Z_C-\log\hat Z_C(k^\ast)\big]^2$
solves the $q$-weighted normal equations
$\mathbb E_{\mathcal Q}\big[(\log\tfrac{Z_C}{n}-q^\top k^\ast/\sqrt d)\,q\big]=0$; it can shrink
the \emph{average} bias but not the per-query one, and forfeits the clean certificate. The
attention-weighted key $k^\ast=\sum_i\bar w_i k_i$ is a cheap heuristic surrogate.

\section{Coupling of the two channels}

\begin{proposition}[Why the channels cannot be optimised independently]\label{prop:joint}
Writing $\hat o-o=\sum_C(\hat p_C-p_C)u_C+\sum_C\hat p_C(v^\ast_C-u_C)$, the first (mass)
and second (value) sums share the normaliser and, under the structured compressing-query
law, are anti-correlated: a mass over-estimate on $C$ inflates $\hat p_C$ exactly where the
value residual $v^\ast_C-u_C$ is systematically signed, partially cancelling. Hence
optimising $v^\ast$ at fixed $k^\ast=\bar k$ is the safe first target; a joint $(k^\ast,v^\ast)$
change must be scored on the full error, with the joint oracle $(Z_C,u_C)$ (zero error) as
ceiling.
\end{proposition}

This matches the earlier estimator-correction finding that a per-query first-order value
fix hurts the total error; Proposition~\ref{prop:value} differs in using a fixed
$\mathcal Q$-averaged reweighting, which is untested and is the object of the offline study.

\section{Predictions to test offline}
\begin{enumerate}
\item $v^\ast=\sum_i\bar w_i v_i$ (compress-distribution weights) reduces the true output
  error over $\bar v$, and closes a measurable fraction of the $\bar v\!\to\!u_C$ gap.
\item Using the compressing-query law $\mathcal Q$ beats an all-query average $\bar w^{\text{all}}$.
\item The key surrogate $k^\ast=\sum_i\bar w_i k_i$ helps the mass estimate marginally but
  loosens the certificate; the anti-correlation of Prop.~\ref{prop:joint} may erase the
  joint gain.
\end{enumerate}

\end{document}
